Computational methods in landscape ecology

### 2024-09-18 Sections/02-spatial-patterns.qmd
Update BIB

<!-- MH: Make sure this includes all similar papers, e.g., Uuemaa2011 -->
To address some limitations, such as correlation and consistency, multivariate factor analysis [@Riitters1995], principal component analysis [@Cushman2008Parsimony], the variance inflation factor [@Plexida2014], or classification trees [@Breiman2017] were used to identify main components of landscape patterns.
<!-- MH: Add one sentence summarising studies above if possible -->
More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
This is in line with various reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency, which are connected to complexity and aggregation as suggested by @Nowosad2018.
<!-- MB: I wonder if we could already mention [@Nowosad2019InformationTheory] here since it presents the same two-component view of landscapes and links with the following section quite naturally -->

Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpen], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
<!-- MB: Are there any libraries to perform factor analysis in R that are kind of specific to landscape ecology? I have had a draft on how to perform factor analysis with pylandstats https://github.com/martibosch/pylandstats-notebooks/blob/main/notebooks/05-cluster-analysis.ipynb, hopefully I will include it in the next release before -->
<!-- MH: Would not mention PCA or similar here. Thats a general statistical issue and outside scope imho -->

@@

R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@gillies2019] in Python, and GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia. 
These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.

>>>>

<!-- MH: Make sure this includes all similar papers, e.g., Uuemaa2011 -->
To address some limitations, such as correlation and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees were used to identify main components of landscape patterns.
<!-- MH: Add one sentence summarising studies above if possible -->
More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
This is in line with various reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency, which are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.
<!-- MB: I wonder if we could already mention [@Nowosad2019InformationTheory] here since it presents the same two-component view of landscapes and links with the following section quite naturally -->

Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpensource], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
<!-- MB: Are there any libraries to perform factor analysis in R that are kind of specific to landscape ecology? I have had a draft on how to perform factor analysis with pylandstats https://github.com/martibosch/pylandstats-notebooks/blob/main/notebooks/05-cluster-analysis.ipynb, hopefully I will include it in the next release before -->
<!-- MH: Would not mention PCA or similar here. Thats a general statistical issue and outside scope imho -->

@@

R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia. 
These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.

### 2024-09-18 Sections/08-discussion.qmd
Update BIB

Since emerging in the 1980s [@Turner2005LandscapeEcology], the field of landscape ecology has constantly evolved [@Costanza2019DescribingAnalyzing; @Gustafson2019HowHas] and is now highly depending on computational methods.
Thus, future progress within the research field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
To maximize the benefits of "Open science", such as shareability, reproducibility, and transparency, the research community will gain the most if both existing and new software support script-based workflows and are made available under open-source licenses [@Powers2019OpenScience; @Vogt2019PatternsSoftware].
<!--JN: maybe add a sentence or two summarizing the previous sections (e.g., that this paper presents various computational methods used in landscape ecology)-->

>>>>

Since emerging in the 1980s <!-- REF -->, the field of landscape ecology has constantly evolved [@Costanza2019DescribingAnalyzing; @Gustafson2019HowHas] and is now highly depending on computational methods.
Thus, future progress within the research field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
To maximize the benefits of "Open science", such as shareability, reproducibility, and transparency, the research community will gain the most if both existing and new software support script-based workflows and are made available under open-source licenses [@Powers2019OpenScience; @Vogt2019PatternsSoftware].
<!--JN: maybe add a sentence or two summarizing the previous sections (e.g., that this paper presents various computational methods used in landscape ecology)-->
